### 2020-05-15 serial.tex
done

\begin{flushleft}{
	\Large
	\textbf\newline{
		Unraveling the paradox between generation and serial intervals
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
Kaiyuan Sun\textsuperscript{2}
Benjamin M.\ Bolker\textsuperscript{3,4,5}
David Champredon\textsuperscript{6}
David J.\,D.\ Earn\textsuperscript{4,5}
Michael Li\textsuperscript{3}
Joshua S.\ Weitz\textsuperscript{7, 8}
Bryan T.\ Grenfell\textsuperscript{1,2,9}
Jonathan Dushoff\textsuperscript{3,4,5,*}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
\\
\textbf{3} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{4} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{5} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\textbf{6} Department of Pathology and Laboratory Medicine, University of Western Ontario, London, Ontario, Canada
\\
\textbf{7} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{8} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{9} Woodrow Wilson School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\bigskip

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.95\textwidth]{forward.pdf}
\caption{
\textbf{Epidemiological dynamics and changes in mean forward and backward delay distributions.}
(A) Daily incidence over time.
(B--D) Changes in the mean forward incubation period, generation interval, and serial interval.
(E--G) Changes in the mean backward incubation period, generation interval, and serial interval.
Black lines represent the results of a deterministic simulation.
Orange lines (A) represent the results of 10 stochastic simulations.
Orange points (B-G) represent the average of 10 stochastic simulations.
\jd{I don't see orange lines, and the location of the dashed line is not explained. I guess you use the forward value at time 0? Would the intrinsic value be better?}
Intrinsic incubation periods and intrinsic generation-intervals are assumed to be independent of each other.
See Table 1 for parameter values.
}
\label{fig:epi}
\end{center}
\end{figure}

The initial forward serial interval distribution captures the exponential growth phase of an epidemic.
Now, we characterize how forward and backward serial intervals vary over the course of an epidemic.
While the forward serial-interval distribution is our primary focus, characterizing the difference between the forward and the backward distributions are important because the observed intervals during on ongoing epidemic are like backward distributions: 
we typically identify infected individuals and ask when they were infected by whom.
Similarly, when we are estimating the incubation period of an individual, we often observe their symptom onset date and try to estimate when they were infected.

@@

The forward serial interval distributions depend on three intervals (\fref{diagram}): (i) the backward incubation period distributions, (ii) the forward generation-interval distributions, and (iii) the forward incubation period distributions.
Since both forward incubation period (\fref{epi}B) and generation-interval (\fref{epi}C) distributions remain roughly constant, changes in the forward serial-interval distributions are predominantly driven by changes in the backward incubation period distributions, whose mean increases over time (\fref{epi}D). \jd{Are you saying that this is true in general, or for these parameters? Have we said enough about these parameters to make this clear to others?}
In this case, the increase in the mean backward incubation period directly translates to the decrease in the mean forward serial interval.
Therefore, the forward serial intervals can decrease over time even in the absence of significant depletion of the susceptible pool;
decrease in incidence via intervention strategies can still cause the backward incubation period to increase, which in turn can decrease serial intervals. 
\jd{Bringing intervention in here is potentially confusing. If you want to do it, maybe try a bit more lead in?} 
In general, relative contributions of the three distributions depend on their shapes, correlations between incubation periods and generation intervals, and overall epidemiological dynamics.

We see similar qualitative patterns in all three backward delays (\fref{epi}E--G; \eref{backward}), because they are predominantly driven by the rate of change in incidence, which in turn affects relative cohort sizes.
When incidence is increasing, individuals are more likely to have been infected recently and therefore, we are more likely to observe shorter intervals (\eref{backexp}).
Similarly, when incidence decreases, we are more likely to observe longer intervals.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{diagram.pdf}
\caption{
\textbf{Estimating the reproduction number from the observed serial intervals.}
(A) Schematic representation of line list data collected during an epidemic.
(B) Estimates of \Ro based on all observed serial intervals completed by a given time. \jd{Is this accurate?}
(C) Schematic representation of line list data rearranged by symptom onset date of infectors.
(D) Estimates of \Ro based on all observed serial intervals started by a given time. 
Black dashed lines represent the mean initial forward serial interval and \Ro.
Black solid lines represent the mean intrinsic serial interval and \Rintrinsic.
Colored solid lines represent the mean estimates of \Ro across 10 stochastic simulations.
Colored ribbons represent the range of estimates of \Ro across 10 stochastic simulations.
}
\label{fig:obsrR}
\end{figure}

Now we turn to practical issues.
In order to have an unbiased estimate of \Ro, we need to use serial-interval distributions estimated based on primary cohorts (i.e., infectors who share the same symptom onset time) at the early stage of the epidemic (i.e. initial forward serial-interval distribution).
However, when an epidemic is ongoing, the observed serial intervals are subject to right-censoring because we cannot observe a serial interval if either an infector or an infectee has not yet developed symptoms;
for example, if we were to measure serial intervals on Day 8 as in \fref{obsrR}A, we will only be able to observe the first 6 events (ID 1--6).
\fref{obsrR}B demonstrates how the effect of right-censoring in the observed serial intervals translates to the underestimation of the basic reproduction number \Ro in our stochastic simulations.
Notably, even if we could observe, and average, \emph{all} serial intervals across all transmission pairs after the epidemic has ended, we would still underestimate the initial mean forward serial interval (and therefore \Ro), likely by a large amount, because the observed serial-interval distribution converges to the intrinsic serial-interval distribution
(in fact, we would even underestimate the intrinsic value slightly due to contraction of the forward generation-interval distribution during the susceptible depletion phase (\fref{epi}C)).

Here we provide a heuristic way of assessing potential biases in the estimate of the mean initial forward serial interval and therefore \Ro retrospectively.
Once serial intervals have been observed after the epidemic has been sufficiently progressed, we can rearrange the line list and group observed serial intervals based on the symptom onset date of infectors (\fref{obsrR}C).
Then, we can compare how estimates of the mean serial interval as well as \Ro change as we incorporate more recent cohorts into the analysis;
that is, we analyze observed serial intervals from infectors who became symptomatic before time $t$ and evaluate how the estimates change as we increase $t$.
This approach is analogous to averaging over a set of forward intervals, just as using all information up to a certain time is analogous to averaging over a set of backward intervals.
(\fref{obsrR}D).
During the exponential growth phase, the estimates of the mean serial interval and \Ro are consistent with the target value;
adding more data allows us to make more precise inference during this period.
However, the cohort-averaged estimates decrease rapidly soon after the exponential growth period, reflecting changes in the forward serial-interval distributions.
This approach allows us to detect dynamical changes in the forward serial-interval distributions and its effect on the estimates of \Ro.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{serial_analysis.pdf}
\caption{
\textbf{Observed serial intervals of COVID-19 and cohort-averaged estimates of \RR.}
(A--B) forward and backward serial intervals over time.
Points represent the means. 
Vertical error bars represent the 95\% equi-tailed quantiles.
Solid lines represent the estimated locally estimated scatterplot smoothing (LOESS) fits.
The dashed line represents the maximum observable delay for each primary cohort calculated from the most recent observed symptom onset date.
(C) Cohort-averaged estimates of \Ro assuming doubling period of 6 and 8 days \citep{li2020early, wu2020nowcasting}.
Ribbons represent the associated 95\% bootstrap confidence intervals.
The data were taken from Supplementary Materials of \cite{du2020serial}.
}
\label{fig:du}
\end{figure}

@@

Finally, we re-analyze serial intervals of COVID-19 collected by \cite{du2020serial} from mainland China, outside Hubei province, based on transmission events reported between January 21--February 8, 2020;
\cite{du2020serial} estimated the mean serial of 3.96 days (95\% CI 3.53–4.39 days) and \Ro of 1.32 (95\% CI 1.16–1.48).
\fref{du}A shows that the mean forward serial interval decreases over time.
While the decrease is likely to be affected by the right-censoring (indicated by the closeness between the quantiles of the observed serial intervals and maximum \emph{observable} serial intervals), increase in the proportion of negative serial intervals are clearly indicative of the changes in the forward serial-interval distribution.
\jd{Is censoring relevant to negative values as well? We might not have (or can't?) observe events before a certain time? It's a bit confusing.}
The decrease in the mean forward serial interval was likely caused by increase in the mean backward serial intervals \jd{incubation periods? is it obvious to the reader why these are increasing?} as well as intervention efforts which reduce generation intervals by preventing infections once cases are identified.
\fref{du}B shows that the mean backward serial interval increases over time, reflecting decrease in incidence over time.

While the qualitative changes in the mean forward and backward serial interval are consistent with our earlier simulations (\fref{epi}), the initial mean forward serial interval (\fref{du}A) appears to be larger than what we calculate earlier based on previously estimated incubation period and generation-interval distributions (\fref{rR}C).
These results indicate that previous estimates of incubation period and generation-intervals may have been underestimated as neither study explicitly accounts for right-censoring (Table 1).

\fref{du}C shows the cohort-averaged estimates of \Ro, which remain roughly constant until day 7 and suddenly decreases;
this sudden decrease is indicative of changes in the forward serial intervals as shown in our simulation study (\fref{obsrR}).
The estimates of \Ro based on the early forward serial intervals are also consistent with previous estimates of \Ro of the COVID-19 epidemic in China \citep{majumder2020early, park2020reconciling}.
We note that early cohort-averaged estimates of \Ro are unlikely to be affected by the right-censoring as we expect the degree of right-censoring to be low (\fref{du}A).
This example demonstrates the danger of using the observed serial intervals to calculate the basic reproduction number without integrating serial intervals into cohorts..

@@

Generation- and serial-interval distributions determine the time scale of disease transmission, and are therefore critical to dynamical modeling of infectious outbreaks.
Here, we show that the initial \emph{forward} serial-interval distribution --- measured during the exponential growth phase of an epidemic --- provides the correct link between the exponential growth rate $r$ and the reproduction number \RR.
In general, this will not match the intrinsic or the backward serial interval.
Further, this interval can decrease over time due to disease dynamics. 
Failing to account for these differences can result in underestimation of \RR.

Our study underlines the importance of carefully defining measured epidemiological time distributions.
Previous studies have shown that generation interval can be either measured forward or backward \citep{nishiura2010time,champredon2015intrinsic,britton2019estimation};
we generalize these ideas and show that they apply to other epidemiological distributions.
Changes in the backward delay distributions due to changing cohort sizes is expected to be a pervasive feature of outbreak dynamics.
Recent analyses of COVID-19 epidemics have attempted to reconstruct epidemic time series of symptomatic cases or infected cases from the reports of confirmed cases by using time-independent delay distributions from infection or symptom onset to reporting (e.g., \cite{tempvar, park2020potential, shim2020transmission});
these studies effectively assume constant backward delay distributions.
Although such approaches may be able to roughly match the time scale of an epidemic, we recommend using deconvolution approaches \citep{goldstein2009reconstructing} or mechanstic modeling approaches for proper reconstruction of epidemic time series instead \citep{flaxman2020estimating}.
\jd{I'm not in love with Goldstein as a practical approach, even though it was more or less my idea! I'd like to learn more about Flaxman.}

While our results support the use of serial interval distributions for calculating \RR, 
they also reveal gaps in current practices in incorporating serial-interval distributions into outbreak analyses.
In particular, \cite{thompson2019improved} recently emphasized the importance of using up-to-date serial interval data for accurate estimation of time-varying reproduction numbers;
however, our study demonstrates that early serial interval data is required to estimate \RR and using up-to-date serial interval data can, in fact, underestimate \RR.
\jd{I'm confused by this. We should be working towards methods that don't get the wrong answer just because you give them more information. Also, if \RR is assumed to be time-varying then of course you want to be up-to-date, no?}
Moreover, \cite{tempvar} recently analyzed case reports of COVID-19 from over 30 countries to estimate their time-varying reproduction numbers but relied on a [fixed] serial-interval distribution;
given large geographical differences in the observed exponential growth rates of the COVID-19 epidemic, the initial forward serial interval distributions is likely to vary across different countries.
\jd{Isn't it more interesting for our purposes whether they appropriately considered the \emph{forward} interval?}
We suggest that modelers should aim to characterize spatiotemporal variation in initial forward serial-interval distributions and understand their changes over time.
These modeling approaches should be necessarily coupled with intensive epidemiological investigation through contact tracing at the very beginning of the outbreak in order to characterize the initial forward serial-interval distribution, particularly for an emerging epidemic such as COVID-19.

Our study also underlines the fact that the serial-interval distribution depends not only on the generation-interval and incubation-period distributions, but also on their correlations.
This implies that not considering this correlation can bias the estimates of the intrinsic generation-interval distribution from the serial-interval distribution.
Furthermore, previous studies that tried to estimate the generation-interval distributions from the observed serial intervals often ignored the differences between the incubation periods of infectors and those of infectees (e.g., \cite{klinkenberg2011correlation, ganyani2020estimating});
\jd{Need to be clearer what source of difference is relevant here.}
There is currently a need for better statistical tools for teasing apart the intrinsic generation-interval distributions from the observed serial intervals.

Our study is not without limitations.
We assumed that all individuals develop symptoms and that the the entire transmission process (e.g., direction of transmission, symptom onset dates, infection dates, etc.) are observable. \jd{I don't think we quite assumed that infection dates are observable, we can discuss.}
In practice, asymptomatic and presymptomatic transmission of COVID-19 makes measuring serial intervals difficult \citep{bai2020presumed,he2020temporal,wei2020presymptomatic}.
For example, symptom onset dates cannot be used as a reliable proxy for determining the direction of transmission as infectees may develop symptoms before their infectors. \jd{Not an example, and not really something we don't consider.}
Based on the COVID-19 parameters (Table 1), we expect roughly 3\% of infectees to develop symptoms before their infectors during the exponential growth period; 
this probability increases as epidemic progresses because the mean backward incubation period increases.
Biases in the observed serial intervals will necessarily bias the estimates of \RR. 
Furthermore, serial intervals do not take into account asymptomatic transmission; 
not explicitly accounting for asymptomatic transmission may also bias the estimates of \RR \citep{park2020time}.

While we focused on the effect of epidemiological dynamics on the serial-interval distributions, other factors, such as intervention strategies, can also affect the generation- and serial-interval distributions.
Individual-level interventions, such as case isolation, directly affect individuals' ability to transmit and will shorten the forward generation- and serial-intervals.
Population-level interventions, such as social distancing, have qualitatively similar effects on epidemiogical dynamics as susceptible depletion;
therefore, we expect these intervention efforts to also (indirectly) shorten the forward generation- and serial-intervals.
\swp{Sunky suggested: ``Social distancing will squeeze the contacts into the household, thus introduce high degree of clustering, which will shorten the generation interval as well (running out of susceptible within the household before the infectious period ends).'' and I like the idea of mentioning this briefly and cite our previous contact\_trace paper... don't have a good sentence at the moment so I'm just going to leave it as is.}
\jd{Maybe leave out the susceptible depletion and just jump straight to the sunky/trace stuff? }

@@

>>>>

\begin{flushleft}{
	\Large
	\textbf\newline{
		Unraveling the paradox between generation and serial intervals
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
Kaiyuan Sun\textsuperscript{2}
Benjamin M.\ Bolker\textsuperscript{3,4,5}
David Champredon\textsuperscript{6}
David J.\,D.\ Earn\textsuperscript{4,5}
Michael Li\textsuperscript{3}
Joshua S.\ Weitz\textsuperscript{7, 8}
Bryan T.\ Grenfell\textsuperscript{1,2,9}
Jonathan Dushoff\textsuperscript{3,4,5}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
\\
\textbf{3} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{4} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{5} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\textbf{6} Department of Pathology and Laboratory Medicine, University of Western Ontario, London, Ontario, Canada
\\
\textbf{7} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{8} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{9} Woodrow Wilson School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\bigskip

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.95\textwidth]{forward.pdf}
\caption{
\textbf{Epidemiological dynamics and changes in mean forward and backward delay distributions.}
(A) Daily incidence over time.
(B--D) Changes in the mean forward incubation period, generation interval, and serial interval.
(E--G) Changes in the mean backward incubation period, generation interval, and serial interval.
Black lines represent the results of a deterministic simulation.
Orange lines (A) represent the results of 10 stochastic simulations.
Orange points (B-G) represent the average of 10 stochastic simulations.
Dashed lines represent the mean initial forward delay.
Intrinsic incubation periods and intrinsic generation-intervals are assumed to be independent of each other.
See Table 1 for parameter values.
}
\label{fig:epi}
\end{center}
\end{figure}

The initial forward serial interval distribution captures the exponential growth phase of an epidemic.
Now, we characterize how forward and backward serial intervals vary over the course of an epidemic.
While the forward serial-interval distribution is our primary focus, understanding the differences between the forward and the backward distributions are important because the observed intervals during on ongoing epidemic are like backward distributions:
we typically identify infected individuals and ask when they were infected by whom.
Similarly, when we are estimating the incubation period of an individual, we often observe their symptom onset date of infected individuals and try to estimate when they were infected (e.g., \cite{backer2020incubation}).

@@

The forward serial interval distributions depend on three intervals (\fref{diagram}): (i) the backward incubation period distributions, (ii) the forward generation-interval distributions, and (iii) the forward incubation period distributions.
For COVID-19 parameters (Table 1), both forward incubation period (\fref{epi}B) and generation-interval (\fref{epi}C) distributions remain roughly constant;
therefore, changes in the forward serial-interval distributions are predominantly driven by changes in the backward incubation period distributions, whose mean increases over time (\fref{epi}D).
In this case, the increase in the mean backward incubation period directly translates to the decrease in the mean forward serial interval.
Since the increase in the mean backward period is driven by the decreasing incidence, which can still occur in the absence of significant depletion of the susceptible pool via intervention, decrease in the mean forward serial interval may be a robust feature in COVID-19 outbreaks.
In general, relative contributions of the three distributions depend on their shapes, correlations between intrinsic incubation periods and generation intervals, and overall epidemiological dynamics.

We see similar qualitative patterns in all three backward delays (\fref{epi}E--G; \eref{backward}), because they are predominantly driven by the rate of change in incidence, which in turn affects relative cohort sizes.
When incidence is increasing, individuals are more likely to have been infected recently and therefore, we are more likely to observe shorter intervals (\eref{backexp}).
Similarly, when incidence decreases, we are more likely to observe longer intervals.
Neglecting these changes in the backward distributions will necessarily bias the inference of the intrinsic distributions from the observed distributions.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{diagram.pdf}
\caption{
\textbf{Estimating the reproduction number from the observed serial intervals.}
(A) Schematic representation of line list data collected during an epidemic.
(B) Estimates of \Ro based on all observed serial intervals completed by a given time.
(C) Schematic representation of line list data rearranged by symptom onset date of infectors.
(D) Estimates of \Ro based on all observed serial intervals started by a given time. 
Black dashed lines represent the mean initial forward serial interval and \Ro.
Black solid lines represent the mean intrinsic serial interval and \Rintrinsic.
Colored solid lines represent the mean estimates of \Ro across 10 stochastic simulations.
Colored ribbons represent the range of estimates of \Ro across 10 stochastic simulations.
}
\label{fig:obsrR}
\end{figure}

Now we turn to practical issues.
In order to have an unbiased estimate of \Ro, we need to use serial-interval distributions estimated based on primary cohorts (i.e., infectors who share the same symptom onset time) at the early stage of the epidemic (i.e., initial forward serial-interval distribution).
However, when an epidemic is ongoing, the observed serial intervals are subject to right-censoring because we cannot observe a serial interval if either an infector or an infectee has not yet developed symptoms;
for example, if we were to measure serial intervals on Day 8 as in \fref{obsrR}A, we will only be able to observe the first 6 events (ID 1--6).
\fref{obsrR}B demonstrates how the effect of right-censoring in the observed serial intervals translates to the underestimation of the basic reproduction number \Ro in our stochastic simulations.
Notably, even if we could observe, and average, \emph{all} serial intervals across all transmission pairs after the epidemic has ended, we would still underestimate the initial mean forward serial interval (and therefore \Ro), likely by a large amount, because the observed serial-interval distribution converges to the intrinsic serial-interval distribution;
in fact, we would even underestimate the intrinsic value slightly due to contraction of the forward generation-interval distribution during the susceptible depletion phase (\fref{epi}C).

Here we provide a heuristic way of assessing potential biases in the estimate of the mean initial forward serial interval and therefore \Ro retrospectively.
Once serial intervals have been observed after the epidemic has been sufficiently progressed, we can rearrange the line list and group observed serial intervals based on the symptom onset date of infectors (\fref{obsrR}C).
Then, we can compare how estimates of the mean serial interval as well as \Ro change as we incorporate more recent cohorts into the analysis;
that is, we analyze observed serial intervals from infectors who became symptomatic before time $t$ and evaluate how the estimates change as we increase $t$.
This approach is analogous to averaging over a set of forward intervals, just as using all information up to a certain time is analogous to averaging over a set of backward intervals (\fref{obsrR}D).
During the exponential growth phase, the estimates of the mean serial interval and \Ro are consistent with the target value;
adding more data allows us to make more precise inference during this period.
However, the cohort-averaged estimates decrease rapidly soon after the exponential growth period, reflecting changes in the forward serial-interval distributions.
This approach allows us to detect dynamical changes in the forward serial-interval distributions and its effect on the estimates of \Ro.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{serial_analysis.pdf}
\caption{
\textbf{Observed serial intervals of COVID-19 and cohort-averaged estimates of \RR.}
(A--B) forward and backward serial intervals over time.
Points represent the means. 
Vertical error bars represent the 95\% equi-tailed quantiles.
Solid lines represent the estimated locally estimated scatterplot smoothing (LOESS) fits.
The dashed line represents the maximum and minimum observable delays across the range of .
(C) Cohort-averaged estimates of \Ro assuming doubling period of 6 and 8 days \citep{li2020early, wu2020nowcasting}.
Ribbons represent the associated 95\% bootstrap confidence intervals.
The data were taken from Supplementary Materials of \cite{du2020serial}.
}
\label{fig:du}
\end{figure}

@@

Finally, we re-analyze serial intervals of COVID-19 collected by \cite{du2020serial} from mainland China, outside Hubei province, based on transmission events reported between January 21--February 8, 2020;
\cite{du2020serial} estimated the mean serial of 3.96 days (95\% CI 3.53–4.39 days) and \Ro of 1.32 (95\% CI 1.16–1.48).
\fref{du}A shows that the mean forward serial interval decreases over time.
While the decrease is likely to be affected by the right-censoring (indicated by the closeness between the quantiles of the observed serial intervals and \emph{maximum observable} serial intervals), increase in the proportion of negative serial intervals are clearly indicative of the changes in the forward serial-interval distribution;
increase in the proportion of negative serial intervals is unlikely to be driven by the left-censoring (indicated by the farness between the quantiles of the observed serial intervals and \emph{minimum observable} serial intervals).
The decrease in the mean forward serial interval was likely to be caused by increase in the mean backward incubation periods, reflecting decrease in COVID-19 cases in China, as well as intervention efforts which reduce generation intervals by preventing infections once cases are identified.
\fref{du}B shows that the mean backward serial interval increases over time, also reflecting decrease in COVID-19 cases in China.

While the qualitative changes in the mean forward and backward serial interval are consistent with our earlier simulations (\fref{epi}), the initial mean forward serial interval (\fref{du}A) appears to be larger than what we calculate earlier based on previously estimated incubation period and generation-interval distributions (\fref{rR}C).
These results indicate that previous estimates of incubation period and generation-intervals (Table 1) may have been underestimated as neither study explicitly accounts for the fact that the observed distributions are like backward distributions.

\fref{du}C shows the cohort-averaged estimates of \Ro, which remain roughly constant until day 7 and suddenly decreases;
this sudden decrease is indicative of changes in the forward serial intervals as shown in our simulation study (\fref{obsrR}).
The estimates of \Ro based on the early forward serial intervals are also consistent with previous estimates of \Ro of the COVID-19 epidemic in China \citep{majumder2020early, park2020reconciling}.
We note that early cohort-averaged estimates of \Ro are unlikely to be affected by the right-censoring as we expect the degree of right-censoring to be low (\fref{du}A).
This example demonstrates the danger of using the observed serial intervals to calculate the basic reproduction number without integrating serial intervals into cohorts.

@@

Generation and serial intervals determine the time scale of disease transmission, and are therefore critical to dynamical modeling of infectious outbreaks.
Here, we show that the initial \emph{forward} serial-interval distribution --- measured during the exponential growth phase of an epidemic --- provides the correct link between the exponential growth rate $r$ and the reproduction number \RR.
In general, this will not match the intrinsic or the backward serial interval.
Further, this interval can decrease over time due to disease dynamics. 
Failing to account for these differences can result in underestimation of \RR.

Our study underlines the importance of carefully defining measured epidemiological time distributions.
Previous studies have shown that generation interval can be either measured forward or backward \citep{nishiura2010time,champredon2015intrinsic,britton2019estimation};
we generalize these ideas and show that they apply to other epidemiological distributions.
Changes in the backward delay distributions due to changing cohort sizes is expected to be a pervasive feature of outbreak dynamics.
Recent analyses of COVID-19 epidemics have attempted to reconstruct epidemic time series of symptomatic cases or infected cases from the reports of confirmed cases by using time-independent delay distributions from infection or symptom onset to reporting (e.g., \cite{tempvar, park2020potential, shim2020transmission});
these studies effectively assume constant backward delay distributions.
Although such approaches may be able to roughly match the time scale of an epidemic, we recommend using dynamical approaches that model incidence as well as reporting delays explicitly (e.g., \citep{flaxman2020estimating}).

While our results support the use of serial interval distributions for calculating \RR, 
they also reveal gaps in current practices in incorporating serial-interval distributions into outbreak analyses.
In particular, \cite{thompson2019improved} recently emphasized the importance of using up-to-date serial interval data for accurate estimation of time-varying reproduction numbers;
however, our study demonstrates that early serial interval data is required to estimate \RR and that using up-to-date serial interval data (without accounting for changes in the forward serial-interval distribution) can, in fact, underestimate \RR.
Many studies have already estimated the (time-varying) reproduction numbers for COVID-19 epidemics using aggregated serial interval data (e.g., \cite{tempvar,du2020serial,pan2020jama,zhang2020evolving});
these studies should re-assess whether they appropriately considered the \emph{forward} serial interval.
Given large geographical differences in the observed exponential growth rates of the COVID-19 epidemic, the initial forward serial interval distributions are also likely to vary across different countries.
We suggest that modelers should aim to characterize spatiotemporal variation in initial forward serial-interval distributions and understand their changes over time.
These modeling approaches should be necessarily coupled with intensive epidemiological investigation through contact tracing at the very beginning of the outbreak in order to characterize the initial forward serial-interval distribution, particularly for an emerging epidemic such as COVID-19.

Our study also underlines the fact that the serial-interval distribution depends not only on the generation-interval and incubation-period distributions, but also on their intrinsic correlations.
This implies that not considering this correlation can bias the estimates of the intrinsic generation-interval distribution from the serial-interval distribution.
Furthermore, previous studies that tried to estimate the generation-interval distributions from the observed serial intervals often ignored the dynamical differences between the realized incubation periods of infectors (backward-looking) and those of infectees (forward looking) (e.g., \cite{klinkenberg2011correlation, ganyani2020estimating}).
There is currently a need for better statistical tools for teasing apart the intrinsic generation-interval distributions from the observed serial intervals.

Our study is not without limitations.
We assumed that all individuals develop symptoms and that the the entire transmission process, including all relevant epidemiological delays, are known exactly.
In practice, asymptomatic and presymptomatic transmission of COVID-19 makes measuring serial intervals difficult \citep{bai2020presumed,he2020temporal,wei2020presymptomatic}.
Biases in the observed serial intervals will necessarily bias the estimates of \RR. 
Furthermore, serial intervals do not take into account asymptomatic transmission; 
not explicitly accounting for asymptomatic transmission may also bias the estimates of \RR \citep{park2020time}.

While we focused on the effect of epidemiological dynamics on the serial-interval distributions, other factors, such as intervention strategies, can also affect the generation- and serial-interval distributions.
Individual-level interventions, such as case isolation, directly affect individuals' ability to transmit and will shorten the forward generation- and serial-intervals.
Population-level interventions, such as social distancing, can squeeze contacts into household and introduce high degree of clustering, which, in turn, will shorten the realized generation intervals due to local depletion of the susceptible pool within households \citep{park2019inferring}.

@@

\subsection{Applications: SEIR model}

Consider a Susceptible-Exposed-Infectious-Recovered model:
\begin{equation}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - \beta S I\\
\frac{\mathrm{d}E}{\mathrm{d}t} &= \beta S I - \sigma E\\
\frac{\mathrm{d}I}{\mathrm{d}t} &= \sigma E - \gamma I \\
\frac{\mathrm{d}R}{\mathrm{d}t} &= \gamma I \\
\end{aligned}
\end{equation}
where $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, and $1/\gamma$ is the mean infectious period.
We further assume that the latent period is equivalent to incubation period; in other words, infected individuals can only transmit after symptom onset.
Then, the generation interval will be always longer than the incubation period.

The joint probability distribution of the intrinsic incubation periods and intrinsic generation intervals for this model can be written as:
\begin{equation}
h(x, \tau) = \begin{cases}
0 & x > \tau\\
\gamma \sigma \exp(-\gamma (\tau-x)-\sigma x) & x \leq \tau
\end{cases}
\end{equation}
Then, the intrinsic generation-interval distribution is given by:
\begin{equation}
\begin{aligned}
g(\tau) &= \int_0^\tau h(x, \tau) \dx\\
&= \frac{\gamma \sigma}{\sigma-\gamma} (\exp(-\gamma \tau) - \exp(-\sigma \tau))
\end{aligned}
\end{equation}
On the other hand, the initial forward serial-interval distribution is given by:
\begin{equation}
\begin{aligned}
f_{0}(\tau) &\propto \int_{-\infty}^{0} \int_{0}^{\tau} \exp(r \pinf) h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&\propto \int_{-\infty}^{0} \int_{0}^{\tau}\exp(r \pinf) \exp(-\gamma \sinf+ \sigma \pinf) \exp(-\sigma(\tau-\sinf)) \, \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&\propto  \exp(-\sigma \tau) \int_{-\infty}^{0} \int_{0}^{\tau} \exp((\sigma-\gamma) \sinf) \exp((r+ \sigma) \pinf)\, \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&\propto  (\exp(-\gamma \tau) - \exp(-\sigma \tau)) \int_{-\infty}^{0} \exp((r+ \sigma) \pinf)\,\mathrm{d}\pinf\\
&\propto  \exp(-\gamma \tau) - \exp(-\sigma \tau)
\end{aligned}
\end{equation}
Therefore, both the intrinsic generation interval and the initial forward serial intervals are identically distributed.
